---
title: "**2022 Year in Books**"
author: "Michael Gillis"
output:
  html_document:
    toc: FALSE
    theme:
      text_align: center
      base_font: "Roboto Condensed"
---

<style type="text/css">

h1.title {
  font-size: 75px;
  color: #3f4c5e;
  text-align: center;
}
h4.author {
  font-size: 30px;
  color: #3f4c5e;
  text-align: center;
}
</style>

```{r include=FALSE}

# load packages
library(tidyverse)

# load data
data <- read.csv("/Users/mgillis/Desktop/Projects/year-in-books/2022/Goodreads 2022.csv")

# create mode function
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

# create colorize function
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}

```

---
Every year I track the books that I read through the website Goodreads. Goodreads is a website that began in 2006, and has since been purchased by Amazon, that allows for a user to select books from their extensive catalog, adding the books to shelves like "Read" and "Want to Read." Users can also rate books using a 5 star rating system.

At the end of every year, Goodreads compiles a "Year in Books" for each user. It displays different data, such as total pages read, total books read, and average rating. I found myself wishing that Goodreads would provide more in this annual summary, which lead me to realize that all of the data I needed to create my own exists within the Goodreads catlogue already.

Without further ado, here is my year in books:

# **Books**

```{r echo=FALSE, fig.show="hold", out.width="50%"}

##############
## Barplots ##
##############

## Barplot by Month

# build barplot
base <- ggplot(data = data, mapping = aes(x = Month.Finished))
by_month <- base + geom_bar(aes(fill = Genre), width = 0.85) +
  # use gradient colors to fill the bars based on different genres
  scale_fill_manual(values = c("#061429","#222f43","#3f4c5e","#5e6b7b","#808c98","#a3aeb7","#c9d1d7","#f0f5f8")) +
  theme(
    aspect.ratio = 9/16,
    text = element_text(family = "Roboto Condensed"),
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 10, angle = 30, vjust = 0.6),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(size = 18, face = "bold", color = "#3f4c5e", vjust = -2),
    axis.title.y = element_text(size = 18, face = "bold", color = "#3f4c5e", vjust = 5),
    plot.margin = margin(1.5,1.5,1.5,1.5, "cm"),
    plot.title = element_text(size = 25, face = "bold", color = "#3f4c5e"),
    legend.title = element_text(size = 15, face = "bold", color = "#3f4c5e"),
    legend.key.size = unit(0.8, "cm"),
    legend.background = element_rect(fill = "#E3E3E3"),
    panel.background = element_rect(fill = "#E3E3E3"),
    plot.background = element_rect(fill = "#F1F1F1", color = "#F1F1F1")
  ) +
  # specify order of the bins on the x-axis
  scale_x_discrete(limits = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")) +
  labs(title = "Books Read by Month", x = "Month Finished", y = "Count")

# display plot
by_month

## Barplot by Genre

# build barplot
base <- ggplot(data, aes(x = fct_rev(fct_infreq(Genre))))
by_genre <- base + geom_bar(aes(fill = Genre), width = 0.85) + coord_flip() +
  # use gradient colors to fill the bars based on different genres
  scale_fill_manual(values = c("#061429","#222f43","#3f4c5e","#5e6b7b","#808c98","#a3aeb7","#c9d1d7","#f0f5f8")) +
  geom_text(aes(label = after_stat(count)), stat = "count", color = "#3f4c5e", hjust = -0.3, fontface = "bold", family = "Roboto Condensed") +
  theme(
    aspect.ratio = 9/16,
    plot.margin = margin(1, 1, 1, 1, "cm"),
    text = element_text(family = "Roboto Condensed"),
    panel.grid = element_blank(),
    axis.text.x = element_text(size = 10, vjust = 0.1),
    axis.text.y = element_text(size = 10, angle = 35, hjust = 0.5),
    axis.title.x = element_text(size = 18, face = "bold", vjust = -0.6, color = "#3f4c5e"),
    axis.title.y = element_text(size = 18, face = "bold", vjust = 1, color = "#3f4c5e"),
    legend.position = "none",
    plot.title = element_text(size = 25, face = "bold", color = "#3f4c5e"),
    panel.background = element_rect(fill = "#E3E3E3"),
    plot.background = element_rect(fill = "#F1F1F1", color = "#F1F1F1")
  ) +
  labs(title = "Books Read by Genre", x = "Genre", y = "Count")

#display plot
by_genre
```

I read a total of **`r nrow(data)`** books this year.\
The genre I read the most of this year was **`r colorize(getmode(data$Genre),"#3f4c5e")`**, with a total of **`r colorize(max(table(data$Genre)),"#3f4c5e")` `r colorize(getmode(data$Genre),"#3f4c5e")`** books read.\
I read the most books in **`r colorize(getmode(data$Month),"#3f4c5e")`**, with a total of **`r colorize(max(table(data$Month)), "#3f4c5e")`** books read that month.

# **Pages**
```{r echo=FALSE, fig.show="hold", out.width="50%"}

## line chart by month

#get a row vector of the months.
months <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")

# get page sums by month into a row vector
page_sums <- c(sum(data[which(data$Month.Finished == "January"), 5]),
               sum(data[which(data$Month.Finished == "February"), 5]),
               sum(data[which(data$Month.Finished == "March"), 5]),
               sum(data[which(data$Month.Finished == "April"), 5]),
               sum(data[which(data$Month.Finished == "May"), 5]),
               sum(data[which(data$Month.Finished == "June"), 5]),
               sum(data[which(data$Month.Finished == "July"), 5]),
               sum(data[which(data$Month.Finished == "August"), 5]),
               sum(data[which(data$Month.Finished == "September"), 5]),
               sum(data[which(data$Month.Finished == "October"), 5]),
               sum(data[which(data$Month.Finished == "November"), 5]),
               sum(data[which(data$Month.Finished == "December"), 5]))

#turn those two columns into a data frame
table <- data.frame(months,page_sums)

# build line chart
base <- ggplot(table, aes(x = months, y = page_sums, group = 1))
linechart <- base + geom_line(stat = "identity", color = "#808c98", linewidth = 1.25) +
  geom_point(stat = "identity", color = "#3f4c5e", size = 2.5) +
  # use gradient colors to fill the bars based on different genres
  theme(
    plot.margin = margin(1, 1, 1, 1, "cm"),
    text = element_text(family = "Roboto Condensed"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "#F1F1F1"),
    axis.text.x = element_text(size = 10, angle = 30, vjust = 0.55),
    axis.text.y = element_text(size = 15, hjust = 0.5),
    axis.title.x = element_text(size = 18, face = "bold", vjust = -0.6, color = "#3f4c5e"),
    axis.title.y = element_text(size = 18, face = "bold", vjust = 1.5, color = "#3f4c5e"),
    legend.position = "none",
    plot.title = element_text(size = 25, face = "bold", color = "#3f4c5e"),
    panel.background = element_rect(fill = "#E3E3E3"),
    plot.background = element_rect(fill = "#F1F1F1", color = "#F1F1F1")
  ) +
  scale_x_discrete(limits = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")) +
  labs(title = "Pages Read by Month", x = "Month", y = "Pages")

  
linechart

## cumulative line chart

#get a row vector of the months.
months <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")

# get page sums by month into a row vector
page_sums <- c(sum(data[which(data$Month.Finished == "January"), 5]),
               sum(data[which(data$Month.Finished == "February"), 5]),
               sum(data[which(data$Month.Finished == "March"), 5]),
               sum(data[which(data$Month.Finished == "April"), 5]),
               sum(data[which(data$Month.Finished == "May"), 5]),
               sum(data[which(data$Month.Finished == "June"), 5]),
               sum(data[which(data$Month.Finished == "July"), 5]),
               sum(data[which(data$Month.Finished == "August"), 5]),
               sum(data[which(data$Month.Finished == "September"), 5]),
               sum(data[which(data$Month.Finished == "October"), 5]),
               sum(data[which(data$Month.Finished == "November"), 5]),
               sum(data[which(data$Month.Finished == "December"), 5]))

# create a row vector that contains a cumulative sum for each month
cum_pages <- cumsum(page_sums)

#turn those two columns into a data frame
table2 <- data.frame(months,cum_pages)

# build line chart
base <- ggplot(table, aes(x = months, y = cum_pages, group = 1))
sum_linechart <- base + geom_line(stat = "identity", color = "#808c98", linewidth = 1.25) +
  geom_point(stat = "identity", color = "#3f4c5e", size = 2.5) +
  # use gradient colors to fill the bars based on different genres
  theme(
    plot.margin = margin(1, 1, 1, 1, "cm"),
    text = element_text(family = "Roboto Condensed"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "#F1F1F1"),
    axis.text.x = element_text(size = 10, angle = 30, vjust = 0.55),
    axis.text.y = element_text(size = 15, hjust = 0.5),
    axis.title.x = element_text(size = 18, face = "bold", vjust = -0.6, color = "#3f4c5e"),
    axis.title.y = element_text(size = 18, face = "bold", vjust = 1, color = "#3f4c5e"),
    legend.position = "none",
    plot.title = element_text(size = 25, face = "bold", color = "#3f4c5e"),
    panel.background = element_rect(fill = "#E3E3E3"),
    plot.background = element_rect(fill = "#F1F1F1", color = "#F1F1F1")
  ) +
  scale_x_discrete(limits = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")) +
  labs(title = "Cumulative Pages Read by Month", x = "Month", y = "Pages")

sum_linechart

```

I read a total of **`r sum(data$Pages)`** this year.\
I read the most pages in **`r colorize(table$month[which.max(table$page_sums)], "#3f4c5e")`** reading a total of **`r colorize(max(table$page_sums), "#3f4c5e")`** pages that month.

# **Reviews**



# **Sources**

https://bookdown.org/yihui/rmarkdown-cookbook/font-color.html
