---
title: "**2022 Year in Books**"
author: "Michael Gillis"
output:
  html_document:
    toc: FALSE
    theme:
      text_align: center
      base_font: "Roboto Condensed"
---

<style type="text/css">

h1.title {
  font-size: 75px;
  color: #3f4c5e;
  text-align: center;
}
h4.author {
  font-size: 30px;
  color: #3f4c5e;
  text-align: center;
}
</style>

```{r include=FALSE}

# load packages
library(tidyverse)
library(knitr)

# set date
date <- "2022"

# load data
data <- read.csv("/Users/mgillis/Desktop/Projects/year-in-books/2022/Goodreads 2022.csv")

# create mode function
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

# create colorize function
colorize <- function(x, color) {
  sprintf("<span style='color: %s;'>%s</span>", color, x)
}

# create inline function to format numbers and conditional values
inline_format <- function(x) {
  if(is.numeric(x)) {
    if(abs(x)-round(x) == 0){
      y <- format(x, big.mark = ",")
    } else {
      y <- format(x, digits = 2, nsmall = 2)
    }
  } else {
    y <- x
  }
  
  x <- colorize(paste0("**",y,"**"), "#3f4c5e")
  
}

knit_hooks$set(inline=inline_format)

```

---
Every year I track the books that I read through the website Goodreads. Goodreads is a website that began in 2006, and has since been purchased by Amazon. It allows for a user to select books from their extensive catalog, adding the books to shelves like "Read" and "Want to Read." Users can also rate books using a 5 star rating system.

At the end of every year, Goodreads compiles a "Year in Books" for each user. It displays different data, such as total pages read, total books read, and average rating. I found myself wishing that Goodreads would provide more in this annual summary, which lead me to realize that all of the data I needed to create my own exists within the Goodreads catalog already.

Without further ado, here is my `r date` year in books:

# **Books**

```{r echo=FALSE, fig.show="hold", out.width="50%"}

## build barplot

# initialize
by_month <- ggplot(data, aes(x = Month.Finished)) +
  # plot the bars
  geom_bar(aes(fill = Genre), width = 0.85) +
  # use gradient colors to fill the bars based on different genres
  scale_fill_manual(values = c("#061429","#222f43","#3f4c5e","#5e6b7b","#808c98","#a3aeb7","#c9d1d7","#f0f5f8")) +
  # specify order of the bins on the x-axis
  scale_x_discrete(limits = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")) +
  # add plot labels
  labs(title = "Books Read by Month", x = "Month Finished", y = "Count") +
  # add theme elements
  theme(
    aspect.ratio = 9/16,
    plot.margin = margin(1.5, 1.5, 1.5, 1.5, "cm"),
    text = element_text(family = "Roboto Condensed"),
    axis.text.x = element_text(size = 10, angle = 30, vjust = 0.6),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(size = 18, face = "bold", color = "#3f4c5e", vjust = -2),
    axis.title.y = element_text(size = 18, face = "bold", color = "#3f4c5e", vjust = 5),
    plot.title = element_text(size = 25, face = "bold", color = "#3f4c5e"),
    legend.title = element_text(size = 15, face = "bold", color = "#3f4c5e"),
    legend.key.size = unit(0.8, "cm"),
    legend.background = element_rect(fill = "#E3E3E3"),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "#E3E3E3"),
    plot.background = element_rect(fill = "#F1F1F1", color = "#F1F1F1")
  )

# display the plot
by_month

## build barplot

#initialize
by_genre <- ggplot(data, aes(x = fct_rev(fct_infreq(Genre)))) +
  # flip the coordinates
  coord_flip() +
  # plot the bars
  geom_bar(aes(fill = Genre), width = 0.85) +
  # add the data labels
  geom_text(aes(label = after_stat(count)), stat = "count", color = "#3f4c5e", hjust = -0.3, fontface = "bold", family = "Roboto Condensed") +
  # use gradient colors to fill the bars based on different genres
  scale_fill_manual(values = c("#061429","#222f43","#3f4c5e","#5e6b7b","#808c98","#a3aeb7","#c9d1d7","#f0f5f8")) +
  # change the limits of the y axis (flipped)
  scale_y_continuous(limits = c(0,30)) +
  # add the plot labels
  labs(title = "Books Read by Genre", x = "Genre", y = "Count") +
  # add theme elements
  theme(
    aspect.ratio = 9/16,
    plot.margin = margin(1, 1, 1, 1, "cm"),
    text = element_text(family = "Roboto Condensed"),
    axis.text.x = element_text(size = 10, vjust = 0.1),
    axis.text.y = element_text(size = 10, angle = 35, hjust = 0.5),
    axis.title.x = element_text(size = 18, face = "bold", vjust = -0.6, color = "#3f4c5e"),
    axis.title.y = element_text(size = 18, face = "bold", vjust = 1, color = "#3f4c5e"),
    legend.position = "none",
    plot.title = element_text(size = 25, face = "bold", color = "#3f4c5e"),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "#E3E3E3"),
    plot.background = element_rect(fill = "#F1F1F1", color = "#F1F1F1")
  )

#display the plot
by_genre

```

I read a total of `r nrow(data)` books this year.\
The genre I read the most of this year was `r getmode(data$Genre)`, with a total of `r max(table(data$Genre))` `r getmode(data$Genre)` books read.\
I read the most books in `r getmode(data$Month)`, with a total of `r max(table(data$Month))` books read that month.

# **Pages**
```{r echo=FALSE, fig.show="hold", out.width="50%"}

#get a row vector of the months.
months <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")

# get page sums by month into a row vector
page_sums <- c(sum(data[which(data$Month.Finished == "January"), 5]),
               sum(data[which(data$Month.Finished == "February"), 5]),
               sum(data[which(data$Month.Finished == "March"), 5]),
               sum(data[which(data$Month.Finished == "April"), 5]),
               sum(data[which(data$Month.Finished == "May"), 5]),
               sum(data[which(data$Month.Finished == "June"), 5]),
               sum(data[which(data$Month.Finished == "July"), 5]),
               sum(data[which(data$Month.Finished == "August"), 5]),
               sum(data[which(data$Month.Finished == "September"), 5]),
               sum(data[which(data$Month.Finished == "October"), 5]),
               sum(data[which(data$Month.Finished == "November"), 5]),
               sum(data[which(data$Month.Finished == "December"), 5]))

#turn those two columns into a data frame
table <- data.frame(months,page_sums)

# build linechart

# initialize
pages_linechart <- ggplot(table, aes(x = months, y = page_sums, group = 1)) +
  # plot the line
  geom_line(stat = "identity", color = "#808c98", linewidth = 1.25) +
  # plot the points
  geom_point(stat = "identity", color = "#3f4c5e", size = 2.5) +
  # specify order of the bins on the x-axis
  scale_x_discrete(limits = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")) +
  # add plot labels
  labs(title = "Pages Read by Month", x = "Month", y = "Pages") +
  # add theme elements
  theme(
    plot.margin = margin(1, 1, 1, 1, "cm"),
    text = element_text(family = "Roboto Condensed"),
    axis.text.x = element_text(size = 10, angle = 30, vjust = 0.55),
    axis.text.y = element_text(size = 15, hjust = 0.5),
    axis.title.x = element_text(size = 18, face = "bold", vjust = -0.6, color = "#3f4c5e"),
    axis.title.y = element_text(size = 18, face = "bold", vjust = 1.5, color = "#3f4c5e"),
    plot.title = element_text(size = 25, face = "bold", color = "#3f4c5e"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "#F1F1F1"),
    panel.background = element_rect(fill = "#E3E3E3"),
    plot.background = element_rect(fill = "#F1F1F1", color = "#F1F1F1")
  )

# display the plot
pages_linechart

# create a row vector that contains a cumulative sum for each month
cum_pages <- cumsum(page_sums)

# turn those two columns into a data frame
table2 <- data.frame(months,cum_pages)

## build linechart

# initialize
sum_linechart <- ggplot(table, aes(x = months, y = cum_pages, group = 1)) +
  # plot the line
  geom_line(stat = "identity", color = "#808c98", linewidth = 1.25) +
  # plot the points
  geom_point(stat = "identity", color = "#3f4c5e", size = 2.5) +
  # specify order of the bins on the x-axis
  scale_x_discrete(limits = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")) +
  # add plot labels
  labs(title = "Cumulative Pages Read by Month", x = "Month", y = "Pages") +
  # add theme elements
  theme(
    plot.margin = margin(1, 1, 1, 1, "cm"),
    text = element_text(family = "Roboto Condensed"),
    axis.text.x = element_text(size = 10, angle = 30, vjust = 0.55),
    axis.text.y = element_text(size = 15, hjust = 0.5),
    axis.title.x = element_text(size = 18, face = "bold", vjust = -0.6, color = "#3f4c5e"),
    axis.title.y = element_text(size = 18, face = "bold", vjust = 1, color = "#3f4c5e"),
    plot.title = element_text(size = 25, face = "bold", color = "#3f4c5e"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_line(color = "#F1F1F1"),
    panel.background = element_rect(fill = "#E3E3E3"),
    plot.background = element_rect(fill = "#F1F1F1", color = "#F1F1F1")
  )

# display the plot
sum_linechart

```

I read a total of `r sum(data$Pages)` pages this year. I read an average of `r round(mean(data$Pages))` pages per book.\
I read the most pages in `r table$month[which.max(table$page_sums)]` reading a total of `r max(table$page_sums)` pages that month.\
The longest book I read this year was `r colorize(data$Author[which(data$Pages == max(data$Pages))], "#0000EE")`'s `r colorize(data$Title[which(data$Pages == max(data$Pages))], "#0000EE")`, which has `r max(data$Pages)` pages.

# **Authors**

```{r echo=FALSE, fig.align = "center"}

# create a data frame that is a frequency table for the authors
authors <- count(data, Author)

# create another data frame, this time excluding any author with a frequency < 2
authors_exc <- (authors[which(authors$n != 1),])

# create a final data frame reordered, an ascending order
authors_exc_reordered <- authors_exc %>% arrange(n)

## build barplot

# initialize
barplot_author <- ggplot(data = authors_exc, aes(x = Author, y = n)) + 
  # flip the coordinates
  coord_flip() +
  # plot the bars
  geom_col(width = 0.85, fill = "#3f4c5e") +
  #add the data labels
  geom_text(aes(label = n), hjust = -0.3, color = "#3f4c5e", fontface = "bold", family = "Roboto Condensed") +
  # change the limits of the y axis (flipped)
  scale_y_continuous(limits = c(0,20)) +
  # change the x axis (flipped) limits to be displayed in decreasing order
  scale_x_discrete(limits = c(authors_exc_reordered[,1])) +
  # add plot labels
  labs(title = "Books Read by Author (Excluding one-offs)", x = "Author", y = "Count") +
  # add theme elements
  theme(
    aspect.ratio = 9/16,
    plot.margin = margin(1, 1, 1, 1, "cm"),
    text = element_text(family = "Roboto Condensed"),
    axis.text.x = element_text(size = 10, vjust = 0.1),
    axis.text.y = element_text(size = 10, angle = 35, hjust = 0.5),
    axis.title.x = element_text(size = 18, face = "bold", vjust = -0.6, color = "#3f4c5e"),
    axis.title.y = element_text(size = 18, face = "bold", vjust = 1, color = "#3f4c5e"),
    plot.title = element_text(size = 21, face = "bold", color = "#3f4c5e"),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "#E3E3E3"),
    plot.background = element_rect(fill = "#F1F1F1", color = "#F1F1F1")
  )

# display the plot
barplot_author

```

The author I read most from this year was `r colorize(authors[which(authors$n == max((authors[which(authors$n != 1),2]))),1], "#0000EE")`. I read a total of `r max((authors[which(authors$n != 1),2]))` of their books.\
There were `r sum(authors$n[authors$n == 1])` "one-off" authors, or authors from which I only read a single book.

# **Ratings**

```{r echo=FALSE, warning=FALSE, fig.show="hold", out.width="50%"}

## build histogram

# initialize
my_reviews <- ggplot(data, aes(x = My.Rating)) +
  # plot the bars
  geom_histogram(binwidth = 1, fill = "#3f4c5e") +
  #add the data labels
  stat_bin(binwidth = 1, geom = "text", aes(label = after_stat(count)), vjust = -0.3, fontface = "bold", family = "Roboto Condensed", color = "#3f4c5e", size = 5) +
  # add plot labels
  labs(title = "Distribution of My Ratings", x = "Rating (5 Star Scale)", y = "Count") +
  # change the limits of the y-axis
  scale_y_continuous(limits = c(0,20)) +
  # add theme elements
  theme(
    aspect.ratio = 9/16,
    plot.margin = margin(1, 1, 1, 1, "cm"),
    text = element_text(family = "Roboto Condensed"),
    axis.text.x = element_text(size = 10, vjust = 0.1),
    axis.text.y = element_text(size = 10, hjust = 0.5),
    axis.title.x = element_text(size = 18, face = "bold", vjust = -0.6, color = "#3f4c5e"),
    axis.title.y = element_text(size = 18, face = "bold", vjust = 1, color = "#3f4c5e"),
    plot.title = element_text(size = 25, face = "bold", color = "#3f4c5e"),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "#E3E3E3"),
    plot.background = element_rect(fill = "#F1F1F1", color = "#F1F1F1")
  )

# display the plot
my_reviews

## build histogram

# initialize
gr_reviews <- ggplot(data, aes(x = GR.Rating)) +
  # plot the bars
  geom_histogram(binwidth = 0.1, fill = "#3f4c5e") +
  #add the data labels
  stat_bin(binwidth = 0.1, geom = "text", aes(label = after_stat(if_else (condition = count > 0, as.character(count), ""))), vjust = -0.3, fontface = "bold", family = "Roboto Condensed", color = "#3f4c5e", size = 5) +
  # add plot labels
  labs(title = "Distribution of Goodreads Ratings", x = "Rating (5 Star Scale)", y = "Count") +
  # change the limits of the x-axis
  scale_x_continuous(limits = c(3,5.1)) +
  # change the limits of the y-axis
  scale_y_continuous(limits = c(0,16)) +
  # add theme elements
  theme(
    aspect.ratio = 9/16,
    plot.margin = margin(1, 1, 1, 1, "cm"),
    text = element_text(family = "Roboto Condensed"),
    axis.text.x = element_text(size = 10, vjust = 0.1),
    axis.text.y = element_text(size = 10, hjust = 0.5),
    axis.title.x = element_text(size = 18, face = "bold", vjust = -0.6, color = "#3f4c5e"),
    axis.title.y = element_text(size = 18, face = "bold", vjust = 1, color = "#3f4c5e"),
    plot.title = element_text(size = 25, face = "bold", color = "#3f4c5e"),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "#E3E3E3"),
    plot.background = element_rect(fill = "#F1F1F1", color = "#F1F1F1")
  )

# display the plot
gr_reviews

```

I reviewed a total of `r length(data$My.Rating[which(data$My.Rating != 0)])` books this year. The average rating that I gave to a book was `r mean(data$My.Rating[which(data$My.Rating != 0)])` `r colorize("Stars", "#3f4c5e")`.\
I gave `r length(which(data$My.Rating == 5))` books 5 Stars this year.\
The average rating from Goodreads users of books that I read this year was `r mean(data$GR.Rating)` `r colorize("Stars", "#3f4c5e")`.

# **Books Read** {.color-tabs .tabset}

## All Books Read
Here are all the books I read in `r date`:

```{r echo=FALSE}

kable(data[,(1:7)], col.names = c("Month Finished","Title","Author","Genre","Pages","My Rating","Goodreads Rating"))

```

## Favorite Books Read
Here are my favorite books from `r date`:

```{r echo=FALSE}

kable(data[which(data$Favorite == "Y"),(1:7)], col.names = c("Month Finished","Title","Author","Genre","Pages","My Rating","Goodreads Rating"), row.names = FALSE)

```

# **Sources**
  
* **Packages:**
    + Tidyverse: https://www.tidyverse.org/
    + Knitr: https://cran.r-project.org/web/packages/knitr/index.html
    
* **Functions:**
    + Getmode: https://www.tutorialspoint.com/r/r_mean_median_mode.htm
    + Colorize: https://bookdown.org/yihui/rmarkdown-cookbook/font-color.html
    + Inline Hook: https://www.youtube.com/watch?v=a2MedFEQTeA

